A Formalization of Typed Lua

### 2015-06-17 doc/papers/paper/paper.tex

The difference between {\em unique}, {\em open} and
{\em fixed} tables is the guarantees that the type system
gives about the reference to a value of that type. A reference
to an {\em unique} table is guaranteed to have no aliases,
so it allows greater flexibility in reinterpreting its type;
a reference to an {\em open} table is guaranteed to have
only {\em closed} aliases (we will explain the meaning of the {\em closed} tag shortly), which still lets the type system
reinterpret the type of a reference, but with more
restrictions; a {\em fixed} table can have any number of {\em fixed} or {\em closed} aliases, so its type cannot change
anymore.

@@

Well-formed table types also do not allow \emph{unique} and
\emph{open} table types to appear in the type of the values.
Our type system currently cannot
track aliases of table fields, so this restriction is
necessary. This limitation of the type system is the reason the {\sc T-Constructor} rule uses the auxiliary function $fix$
(in the definition of $vt$)
 to change any {\em open} and {\em unique} table types used in the field initializer to {\em fixed}. As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
has type $\{``x":1, ``y":\{``z":2\}_{fixed}\}_{unique}$.

@@

The type system needs to track aliasing of table references,
as the flexibility of {\em unique} and {\em open} tables
depend on respectively having no aliases and having only {\em closed} aliases. The system conservatively assumes that
a table has been aliased if its identifier appears in r-value position, with a couple of exceptions which we will explain shortly.

@@

Lua has first-class functions, so another way of aliasing
an {\em unique} and {\em open} references is by closing
over them in a function body. In the environment used for
typing the function body, all {\em unique} and {\em open}
tables are {\em closed} with the auxiliary function $closeall$.
This is not enough, as assignment
to a closed-over variable can be unsafe. In the following
example, assume that $g$ receives a parameter of type
$\{ ``x": \String \cup \Nil\}_{closed}$:
\begin{center}
	\begin{tabular}{llll}
		\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{open}
			= \{\} \; \mathbf{in}$}\\
		& \multicolumn{3}{l}{$\mathbf{local} \; f:\Void \rightarrow \Void =$}\\
		& & \multicolumn{2}{l}{$\mathbf{fun} \; ():\Void$}\\
		& & & \multicolumn{1}{l}{$a = \{[``x"]=5\} $}\\
		& \multicolumn{3}{l}{$\mathbf{in} \; \lfloor g({<}\{``x":
			 \String \cup \Nil\}_{closed}{>} \; a) \rfloor_{0}$}
	\end{tabular}
\end{center}

@@

In the first idiom, records and objects are
defined through assignment to an initially empty
table. In Typed Lua's type system the type of a
table can change, either by assignment
or by using the table in a context that expects
a different (but compatible) type from its current one.
The system tracks aliasing of table references
in order to do this type evolution in a type-safe
way.

>>>>

The difference between {\em unique}, {\em open} and
{\em fixed} tables is the guarantees that the type system
gives about the reference to a value of that type. A
reference to an {\em unique} table is guaranteed to
point to a table that has no other references to it,
so it allows greater flexibility in reinterpreting its type;
a reference to an {\em open} table is guaranteed to have
only {\em closed} references pointing to the
same table (we will explain the meaning of the {\em closed} tag shortly), which still lets the type system
reinterpret the type of a reference, but with more
restrictions; a reference to a {\em fixed} table can have any number of
{\em fixed} or {\em closed} references point to it, 
so its type cannot change
anymore.

@@

Well-formed table types also do not allow \emph{unique} and
\emph{open} table types to appear in the type of the values.
At the moment, local variables are the only kind of reference that the system
can track, so this restriction is
necessary. This limitation of the type system is the reason the {\sc T-Constructor} rule uses the auxiliary function $fix$
(in the definition of $vt$)
 to change any {\em open} and {\em unique} table types used in the field initializer to {\em fixed}. As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
has type $\{``x":1, ``y":\{``z":2\}_{fixed}\}_{unique}$.

@@

The type system needs to track aliasing of tables,
as the flexibility of {\em unique} and {\em open} tables
depend on respectively having no aliases and having only {\em closed} aliases. The system conservatively assumes that
a table has been aliased if its identifier appears in r-value position, with a couple of exceptions which we will explain shortly.

@@

Lua has first-class functions, so another way of aliasing
an {\em unique} and {\em open} references is by closing
over them in a function body. In the environment used for
typing the function body, all {\em unique} and {\em open}
tables are {\em closed} with the auxiliary function $closeall$.
Assignment to a closed-over table can be unsafe, because the {\em reference} to it is aliased as well. In the following
example, assume that $g$ receives a parameter of type
$\{ ``x": \String \cup \Nil\}_{closed}$:
\begin{center}
	\begin{tabular}{llll}
		\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{open}
			= \{\} \; \mathbf{in}$}\\
		& \multicolumn{3}{l}{$\mathbf{local} \; f:\Void \rightarrow \Void =$}\\
		& & \multicolumn{2}{l}{$\mathbf{fun} \; ():\Void\;\;\;
		 a = \{[``x"]=5\} $}\\
		& \multicolumn{3}{l}{$\mathbf{in} \; \lfloor g({<}\{``x":
			 \String \cup \Nil\}_{closed}{>} \; a) \rfloor_{0}$}
	\end{tabular}
\end{center}

@@

In the first idiom, records and objects are
defined through assignment to an initially empty
table. In Typed Lua's type system the type of a
table can change, either by assignment
or by using the table in a context that expects
a different (but compatible) type from its current one.
The system tracks aliasing of tables
in order to do this type evolution in a type-safe
way.
